A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-02-04 papers/contact/paper.tex

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  Each of these terms has some dependence on the
inhomogeneity in the density.  The ideal gas is purely local, and in
fact must be the only purely local term in the free energy in order
for the functional to satisfy the contact-value theorem.  The
hard-sphere contribution contribution to the free energy has been
thoroughly studied, and is well approximated by the White Bear version
of the Fundamental-Measure Theory (FMT)
functional\cite{roth2002whitebear}.  A functional for the chain
contribution is has recently been introduced\cite{bryk2006density,
  gross2009density}, but is not needed for small associating molecules
such as water.  The dispersion term is long-range and thus has
significant dependence on the density distribution, but because its
relatively weak position dependence, is amenable to mean-field
approximations.  A functional for the association term of the free
energy was introduced by Wu and
Yu\cite{yu2002fmt-dft-inhomogeneous-associating}, and was used in
several subsequent functionals\cite{fu2005vapor-liquid-dft,
  bryk2006density}.  A key input (FIXME: term, factor?) to this
functional is the contact value of the correlation function of a
hard-sphere fluid.

The contact value of the correlation function---also known as the
cavity correlation function---is the the value of the
correlation function evaluated a distance equal to the hard-sphere
diameter.  Yu and Wu introduce an approximation based on the
fundamental measures of Fundamental Measure Theory (see
Section~\ref{sec:yuwu}).  In this paper, we will report on the
accuracy of this approximation by comparing with Monte Carlo
simulations of the inhomogeneous hard-sphere fluid.  In addition, we
will introduce and test differing approximations for the cavity
correlation function.

>>>>

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  Each of these terms has some dependence on the
inhomogeneity in the density.  The ideal gas is purely local, and in
fact must be the only purely local term in the free energy in order
for the functional to satisfy the contact-value theorem.  The
hard-sphere contribution contribution to the free energy has been
thoroughly studied, and is well approximated by the White Bear version
of the Fundamental-Measure Theory (FMT)
functional\cite{roth2002whitebear}.  A functional for the chain
contribution is has recently been introduced\cite{bryk2006density,
  gross2009density}, but is not needed for small associating molecules
such as water.  The dispersion term is long-range and thus has
significant dependence on the density distribution, but because its
relatively weak position dependence, is amenable to mean-field
approximations.  A functional for the association term of the free
energy was introduced by Wu and
Yu\cite{yu2002fmt-dft-inhomogeneous-associating}, and was used in
several subsequent functionals\cite{fu2005vapor-liquid-dft,
  bryk2006density}.  A key input (FIXME: term, factor?) to both the
association and chain free energies is the contact value of the
correlation function of a hard-sphere fluid.

The contact value of the correlation function---which is the same as
the contact value of the cavity correlation function for hard
spheres---is the the value of the correlation function evaluated a
distance equal to the hard-sphere diameter.  In an inhomogeneous
system treated using fundamental measures, there is some ambiguity in
the meaning of the \emph{local} contact value of the correlation
function which appears in the equations for the homogeneous system.
Clearly the contact value of the correlation function will depend on
position, but how should that position be defined? It will depend on
how the correlation function is used.  There are two reasonable
choices for functionals based on fundamental measure theory, which
will depend on which densities are used...
\begin{align}
  F[n(\rr)] = \int n(\rr) f(\bar{n}(\rr), g_\sigma(\rr)) d\rr
\end{align}
\begin{align}
  F[n(\rr)] = \int f(n_2(\rr), g_\sigma(\rr)) d\rr
\end{align}

Yu and Wu introduce an approximation based on the fundamental measures
of Fundamental Measure Theory (see Section~\ref{sec:yuwu}).  Gross
also does\cite{gross2009density}.

In this paper, we will report on the accuracy of this approximation by
comparing with Monte Carlo simulations of the inhomogeneous
hard-sphere fluid.  In addition, we will introduce and test differing
approximations for the cavity correlation function.
